A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-03-05 response/response.tex

{\color{red}  
\begin{itemize}
     \item
 Manuscript corrected.
  \end{itemize}}

 \item Pgs. 21-23: You may capitalize the ﬁrst letters on the journal names in your refer-
ences [9, 20]

{\color{red}  
\begin{itemize}
     \item
 Manuscript corrected.
  \end{itemize}}

 \item Pg. 22: In your reference [10], you may replace “fr{\'e}chet” by “Fr{\'e}chet”.

 \item Pg. 23: Update your reference [21] (see the reference [3] in this report). Also,
update your reference [27].

@@

 \item Pg. 23: Update your reference [21] (see the reference [3] in this report). Also,
update your reference [27].

{\color{red}  
\begin{itemize}
     \item
 Manuscript corrected.
  \end{itemize}}

 \item Pg. 24: In line 18, replace the comma by a period.

 \item Pg. 24: I would replace the comma by a period in line 28, and replace “using
L’H{\ˆo}pital’s rule” by “Using l’H{\ˆo}pital’s rule” in line 31.

### 2014-03-05 jacobian_calc_CMAME.tex
all revisions addressed except peridynamics section

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Frechet derivatives of matrix to matrix functions
\cite{al2010complex}. To compute the Jacobian of the residual in the
peridynamics simulation as discussed in the instant work is analogous is
computing a Frechet derivative. Where the application of the CS derivatives in
the prior work is limited to a particular class of abstract functions to
determine the CS method's suitability for use in condition number estimation, the
instant work applies CS to solving a solid mechanics problem. The instant work is not
about computing Frechet derivatives with CS but moreover using them in a
numerical method to solve a set of equations describing a physical system.

@@

\begin{equation} 
    \label{eqn:TrussForce}
    f(D) = .5AE(cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}cos^{2}\phi - 3sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})sin^{2}\phi,
\end{equation}

\begin{equation} 
    \label{eqn:TrussStiffness}
    S_{T}(D) = 1.5AE(cos^{2}\phi)[\frac{D}{L_{0}}cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AEsin^{2}\phi}{L_{0}}.
\end{equation}

@@

\note[Michael]{re: I.2} \add[Michael]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Frechet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Frechet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

@@

\appendix
\renewcommand*{\thesection}{\Alph{section}}
%
\section{Complex-Step Example}
\label{sec:appendixA}
%
Using $f(x) = \cos(x)$, we have
%
\[
f (x + i h) = \cos(x + i h),
\]
%
or through trigonometric identities equivalently
%
\[
f(x + i h) = \cos(x) \cosh(h) - i \sin(x) \sinh(h),
\]
%
Applying equation (\ref{eqn:complexFirstDeriv})
\[
    \frac{\partial f(x)}{\partial x} = - \frac{\sin(x) \sinh(h)}{h} + \mathcal{O}(h^2).
\]
Since $h$ is arbitrary we desire it to be as small as possible or
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \sinh(h)}{h},
\]
using L'H\^opital's rule
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \cosh(h)}{1},
\]
of course, as $h \to 0$ then $\cosh(h) \to 1$ thereby recovering the exact derivative
\[
\frac{\partial f(x)}{\partial x} =-\sin(x).
\]

\section{AD Example}
\label{sec:appendixB}

\begin{enumerate}
%Set up example problem. The example is meant to walk the reader through the forward AD algorithm
%while encouraging them to rely on their understanding of the chain rule
\item Take an example composition function $f(x) = (sin(cos(x)))^2$\\

@@

\begin{enumerate}
%This first block of subitems is meant to establish that the values of x that the user cares about
%about evaluating functions for are included within the set of numbers that functions g,h,k are allowed 
%to deal with, and that nested functions can take each other as input, where s may not necessarily equal
%x.
\item Given that: \\
\label{given}
\begin{enumerate}
\item $S \in R^1$ 
\item $X \in S$ 
\item $g, h, k \in H: S \rightarrow S$ 
\item $\forall s \in S : g(s) = {s}^2$, $h(s) = sin(s)$, $k(s) = cos(s)$ 
\item
\label{composition}
$\forall x \in X : f(x) = g(h(k(x)))$ \\
\end{enumerate}

@@

\item
%AD doesn't return symbolic expressions for derivatives that can be re-evaluated.
%Instead, generic versions of functions and evaluations of their partials are defined together.
particular values can be identified: \\
$s = s_0,..,s_n,...s_{\infty} \mid n=[0,\infty)$ \\ 
and similarly for the other variables $a,b,x$ \\
\item
functions $u,v,w$ and their partial derivatives w.r.t $s$ are defined such that: \\
\begin{tabular}{l}
%\multicolumn{2}{c}{for $a, b, s$ equal to $a_n, b_n, s_n$:} \\ \hline
for $a, b, s$ equal to $a_n, b_n, s_n$: \\ \hline
$u \mid _{a_n, b_n, s_n} = a_n \cdot s_n^{b_n}$ \\
$\frac{\partial{u}}{\partial{s}} \mid _{a_n, b_n, s_n} = a_n \cdot b_n \cdot s_n^{b_n - 1}$ \\
\\
$v \mid _{a_n, b_n, s_n} = a_n \cdot sin(b_n \cdot s_n)$ \\
$\frac{\partial{v}}{\partial{s}} \mid _{a_n, b_n, s_n} = a_n \cdot b_n \cdot cos(b_n \cdot s_n)$ \\ 
\\
$w \mid _{a_n, b_n, s_n} = a_n \cdot cos(b_n \cdot s_n)$ \\
$\frac{\partial{w}}{\partial{s}} \mid _{a_n, b_n, s_n} = -a_n \cdot b_n \cdot sin(b_n \cdot s_n)$\\
\end{tabular} 
\end{enumerate}
%a comment
\item 
\label{abcvalues}
%General definitions of functions are specialized by arguments. 
Given that it is possible to describe $f(x) = (sin(cos(x)))^2$ in terms the computer understands by inputting
$f(x)$ such that the computer stores an equivalent statement $f(x) = u(a, b, s) \mid_{arguments}$,
iff the arguments of $u,v,w$ are chosen such that $u,v,w$ approximate $g,h,k$ as follows:\\
\begin{tabular}{l l l l | c}
	Function & a & b & s & Approximates Function\\ \hline 
	$u$ & $1$ & $2$ & $v$ & $g$\\ \hline 
	$v$ & $1$ & $1$ & $w$ & $h$\\ \hline 
	$w$ & $1$ & $1$ & $x$ & $k$\\ \hline 	
\end{tabular}
\end{enumerate}

\item 
\label{differentiation}
%AD is equivalent to the chain rule, when the computer has definitions for functions and partials
%that can be adapted to match the composition function.
It follows from \ref{composition} that we can evaluate $\frac{d}{d x}f(x) \mid_{x_0}$ with the chain rule: \\ \\
$\frac{d}{d x}f(x) = \frac{d}{d x} \cdot g(h(k(x))) \mid_{x_0}$ \\
$\frac{d}{d x}f(x) =  \frac{d{g}}{d{h}} \cdot \frac{d{h}}{d{k}}
\cdot \frac{d{k}}{d{x}}\mid_{x_0}$ \\ \\
From the rest of \ref{given} it follows that we can approximate  $\frac{d}{d x}f(x) \mid_{x_0}$ by
specializing the computer's general forms of $u, v, w$ 
according to ~\ref{abcvalues}, with parameters $a, b$ chosen for each function and held as constant, and
evaluating, such that the total derivative of our original function w.r.t $x$ where $x = x_0$ is approximated
by the partial derivative of our equivalent statement, $f(x) = u(a, b, s) \mid_{arguments}$,  w.r.t $s$ 
where $s = x_0$. \\

\end{enumerate}
\label{doingthework}
%This shows how information flows through the nested functions in one expression, like a table of contents
For completeness we write out the computer's steps to evaluate \ref{differentiation} under the conditions of
\ref{abcvalues} with a particular value of $s = x_0$, in equation format: \\ \\
$\frac{\partial}{\partial x}f(x) \mid _{x_0} = \frac{\partial{u}}{\partial{v}} \mid _{1, 2, v \mid _{1, 1, w \mid _{ 1, 1, x_0}}} \cdot \frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0} \cdot \frac{ds}{dx}$\\ \\
%Repeating above in tabular format: \\ \\
%This shows the same information but is ordered to show progression, like chapters
%\begin{tabular}{l l l}
%Current Evaluation & $s$ Value & Partial Derivative \\ \hline
%$w(a, b, s)\mid_{1, 1, s}$ & $x_0$ & $\frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0}$ \\
%$v(a, b, s)\mid_{1, 1, s}$ & $w(a, b, s)\mid_{1, 1, x_0}$ & $\frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0}$ \\
%$u(a, b, s)\mid_{1, 2, s}$ & $v(a, b, s)\mid_{1, 1, w(a, b, s)\mid_{1, 1, x_0}}$ & $\frac{\partial{u}}{\partial{v}} \mid _{1, 2, v \mid _{1, 1, w \mid _{ 1, 1, x_0}}} \cdot \frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0} \cdot 1$ \\
%\end{tabular} \\ \\
%This explains what the equation and table above mean.
Because the computer knew the analytical forms of the partial derivatives of each of $u,v,w$ beforehand, which of course needed to be specified using an AD data structure from a library like Trilinos::Sacado \cite{trilinos} or ADIFOR \cite{bischof1995adifor},
all it needed to do (for 'backward' AD) was:
\begin{enumerate}
\item to evaluate each of $u,v,w$ according to \ref{abcvalues}, in order from $w \rightarrow
v \rightarrow u$, 
\item to remember the values for $x_0$ and the output of each function evaluation besides  $u$, 
\item then to use $x_0$ and the output of the function evaluations as input for the corresponding partial
derivative function  evaluations, and
\item store the individual partials.
\end{enumerate}
Lastly, to compute the partial derivative of the entire composition function, the computer multiplies
the individual partials together in observance of the chain-rule.

>>>>

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Fr\'{e}chet derivatives of matrix to matrix functions
\cite{al2010complex}. To compute the Jacobian of the residual in the
peridynamics simulation as discussed in the instant work is analogous is
computing a Fr\'{e}chet derivative. Where the application of the CS derivatives in
the prior work is limited to a particular class of abstract functions to
determine the CS method's suitability for use in condition number estimation, the
instant work applies CS to solving a solid mechanics problem. The instant work is not
about computing Fr\'{e}chet derivatives with CS but moreover using them in a
numerical method to solve a set of equations describing a physical system.

@@

\begin{equation} 
    \label{eqn:TrussForce}
    f(D) = .5AE(\cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}\cos^{2}\phi - 3\sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})\sin^{2}\phi,
\end{equation}

\begin{equation} 
    \label{eqn:TrussStiffness}
    S_{T}(D) = 1.5AE(\cos^{2}\phi)[\frac{D}{L_{0}}\cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AE\sin^{2}\phi}{L_{0}}.
\end{equation}

@@

\note[Michael]{re: I.2} \add[Michael]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simulation}
The tangent-stiffness matrix is the Fr\'{e}chet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Fr\'{e}chet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

@@

\appendix
\renewcommand*{\thesection}{\Alph{section}}
%
\section{Complex-Step Example}
\label{sec:appendixA}
%
Using $f(x) = \cos(x)$, we have
%
\[
f (x + i h) = \cos(x + i h),
\]
%
or through trigonometric identities equivalently
%
\[
f(x + i h) = \cos(x) \cosh(h) - i \sin(x) \sinh(h).
\]
%
Applying equation (\ref{eqn:complexFirstDeriv})
\[
    \frac{\partial f(x)}{\partial x} = - \frac{\sin(x) \sinh(h)}{h} + \mathcal{O}(h^2).
\]
Since $h$ is arbitrary we desire it to be as small as possible or
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \sinh(h)}{h},
\]
using l'H\^opital's rule
\[
\frac{\partial f(x)}{\partial x} = \lim_{h \to 0} - \frac{\sin(x) \cosh(h)}{1}.
\]
As $h \to 0$ then $\cosh(h) \to 1$ thereby recovering the exact derivative
\[
\frac{\partial f(x)}{\partial x} =-\sin(x).
\]

\section{Automatic differentiation example}
\label{sec:appendixB}

\begin{enumerate}
%Set up example problem. The example is meant to walk the reader through the forward AD algorithm
%while encouraging them to rely on their understanding of the chain rule
\item Take an example composition function $f(x) = (\sin(\cos(x)))^2$\\

@@

\begin{enumerate}
%This first block of subitems is meant to establish that the values of x that the user cares about
%about evaluating functions for are included within the set of numbers that functions g,h,k are allowed 
%to deal with, and that nested functions can take each other as input, where s may not necessarily equal
%x.
\item Given that: \\
\label{given}
\begin{enumerate}
\item $S \in R^1$ 
\item $X \in S$ 
\item $g, h, k \in H: S \rightarrow S$ 
\item $\forall s \in S : g(s) = {s}^2$, $h(s) = \sin(s)$, $k(s) = \cos(s)$ 
\item
\label{composition}
$\forall x \in X : f(x) = g(h(k(x)))$ \\
\end{enumerate}

@@

\item
%AD doesn't return symbolic expressions for derivatives that can be re-evaluated.
%Instead, generic versions of functions and evaluations of their partials are defined together.
particular values can be identified: \\
$s = s_0,..,s_n,...s_{\infty} \mid n=[0,\infty)$ \\ 
and similarly for the other variables $a,b,x$ \\
\item
functions $u,v,w$ and their partial derivatives w.r.t $s$ are defined such that: \\
\begin{tabular}{l}
%\multicolumn{2}{c}{for $a, b, s$ equal to $a_n, b_n, s_n$:} \\ \hline
for $a, b, s$ equal to $a_n, b_n, s_n$: \\ \hline
$u \mid _{a_n, b_n, s_n} = a_n \cdot s_n^{b_n}$ \\
$\frac{\partial{u}}{\partial{s}} \mid _{a_n, b_n, s_n} = a_n \cdot b_n \cdot s_n^{b_n - 1}$ \\
\\
$v \mid _{a_n, b_n, s_n} = a_n \cdot sin(b_n \cdot s_n)$ \\
$\frac{\partial{v}}{\partial{s}} \mid _{a_n, b_n, s_n} = a_n \cdot b_n \cdot cos(b_n \cdot s_n)$ \\ 
\\
$w \mid _{a_n, b_n, s_n} = a_n \cdot cos(b_n \cdot s_n)$ \\
$\frac{\partial{w}}{\partial{s}} \mid _{a_n, b_n, s_n} = -a_n \cdot b_n \cdot sin(b_n \cdot s_n)$\\
\end{tabular} 
\end{enumerate}
%a comment
\item 
\label{abcvalues}
%General definitions of functions are specialized by arguments. 
Given that it is possible to describe $f(x) = (\sin(\cos(x)))^2$ in terms the computer understands by inputting
$f(x)$ such that the computer stores an equivalent statement $f(x) = u(a, b, s) \mid_{arguments}$,
iff the arguments of $u,v,w$ are chosen such that $u,v,w$ approximate $g,h,k$ as follows:\\
\begin{tabular}{l l l l | c}
	Function & a & b & s & Approximates Function\\ \hline 
	$u$ & $1$ & $2$ & $v$ & $g$\\ \hline 
	$v$ & $1$ & $1$ & $w$ & $h$\\ \hline 
	$w$ & $1$ & $1$ & $x$ & $k$\\ \hline 	
\end{tabular}
\end{enumerate}

\item 
\label{differentiation}
%AD is equivalent to the chain rule, when the computer has definitions for functions and partials
%that can be adapted to match the composition function.
It follows from \ref{composition} that we can evaluate $\frac{d}{d x}f(x) \mid_{x_0}$ with the chain rule: \\ \\
$\frac{d}{d x}f(x) = \frac{d}{d x} g(h(k(x))) \mid_{x_0}$ \\
$\frac{d}{d x}f(x) =  \frac{d{g}}{d{h}} \cdot \frac{d{h}}{d{k}}
\cdot \frac{d{k}}{d{x}}\mid_{x_0}$ \\ \\
From the rest of \ref{given} it follows that we can approximate  $\frac{d}{d x}f(x) \mid_{x_0}$ by
specializing the computer's general forms of $u, v, w$ 
according to~\ref{abcvalues}, with parameters $a, b$ chosen for each function as in~\ref{abcvalues} and held constant, and
evaluating, such that the total derivative of our original function w.r.t $x$ where $x = x_0$ is approximated
by the partial derivative of our equivalent statement, $f(x) = u(a, b, s) \mid_{arguments}$,  w.r.t $s$ 
where $s = x_0$. \\

\end{enumerate}
\label{doingthework}
%This shows how information flows through the nested functions in one expression, like a table of contents
For completeness we write out the computer's steps to evaluate \ref{differentiation} under the conditions of
\ref{abcvalues} with a particular value of $s = x_0$, in equation format: \\ \\
$\frac{\partial}{\partial x}f(x) \mid _{x_0} = \frac{\partial{u}}{\partial{v}} \mid _{1, 2, v \mid _{1, 1, w \mid _{ 1, 1, x_0}}} \cdot \frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0} \cdot \frac{ds}{dx}.$\\ \\
%Repeating above in tabular format: \\ \\
%This shows the same information but is ordered to show progression, like chapters
%\begin{tabular}{l l l}
%Current Evaluation & $s$ Value & Partial Derivative \\ \hline
%$w(a, b, s)\mid_{1, 1, s}$ & $x_0$ & $\frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0}$ \\
%$v(a, b, s)\mid_{1, 1, s}$ & $w(a, b, s)\mid_{1, 1, x_0}$ & $\frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0}$ \\
%$u(a, b, s)\mid_{1, 2, s}$ & $v(a, b, s)\mid_{1, 1, w(a, b, s)\mid_{1, 1, x_0}}$ & $\frac{\partial{u}}{\partial{v}} \mid _{1, 2, v \mid _{1, 1, w \mid _{ 1, 1, x_0}}} \cdot \frac{\partial{v}}{\partial{w}} \mid _{1,1, w \mid _{1, 1, x_0}} \cdot \frac{\partial{w}}{\partial{s}} \mid _{1, 1, x_0} \cdot 1$ \\
%\end{tabular} \\ \\
%This explains what the equation and table above mean.
Because the computer knew the analytical forms of the partial derivatives of each of $u,v,w$ beforehand, which of course needed to be specified using an AD data structure from a library like Trilinos::Sacado \cite{trilinos} or ADIFOR \cite{bischof1995adifor},
all it needed to do (for 'backward' AD) was:
\begin{enumerate}
\item to evaluate each of $u,v,w$ according to \ref{abcvalues}, in order from $w \rightarrow
v \rightarrow u$, 
\item to remember the values for $x_0$ and the output of each function evaluation besides  $u$, 
\item then to use $x_0$ and the output of the function evaluations as input for the corresponding partial
derivative function  evaluations, and
\item store the individual partials.
\end{enumerate}
Lastly, to compute the partial derivative of the entire composition function, the computer multiplies
the individual partials together in observance of the chain-rule.
